Figures to insert in text

### 2018-03-15 Chapters/Chapter02.tex
Fix hard location (removing hard-location and hardlocation).

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. This model may be applied in all sort of applications because of its incredible ability to closely reflect the human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where one's fear come from. One just feels it. We may interpret this situation as clues of the present --- a dark alley; a giant metropolitan area; people going on about their lives mostly indifferent from each other; constrained routes ahead and behind you; a general lack of activity; that very feeling that something isn't right... without knowing precisely what isn't right, or even what `right' means... etc. --- recalling past experiences from memory and thus generating the feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, one's brain involuntarily makes an analogy and recognizes the possibility of danger. This flexibility into mapping one situation in another is an important human feature which is hard to replicate in computers.

SDM has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. \citet{Linhares2011} has showed that SDM respects the limits of short-term memory discussed by \citet{Miller1995} and \citet{Cowan2011}. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though intriguing results have been achieved using SDM, it requires counter-productive, duplicate, effort from researchers to build on top of previous work.

@@

Our motivation was our own effort to run our models. As there is no reference implementation, we had to implement our own and run several simulations to ensure that our implementation was correct and bug free. Thus, we had to deviate from our main goal --- which was to test our hypothesis and explore the `idea space' --- and to focus in the implementation itself. Furthermore, new members in our research group had to go through different source codes developed by former members.

@@

\begin{tabular}{cp{\textwidth}}
  $n$ & Number of dimensions, i.e., $n=1,000$. \\
  $N$ & Size of the binary space, $|\{0, 1\}^n| = 2^n$. \\
  $N'$ & Number of hard-locations samples from $\{0, 1\}^n$. Its typical value is 1,000,000, as suggested by \citet{Kanerva1988}. \\
  $H$ & Same as $N'$. \\
  $r$ & Access radius, i.e., when $n=1,000$ and $N'=1,000,000$, its typical value is $451$. This value is calculated to activate, on average, one thousandth of $N'$. \\
  $\eta$ & A bitstring, usually a datum. \\
  $\eta_x$ & A clue $x$ bits away from $\eta$, i.e., $\text{dist}(\eta, \eta_x) = x$. \\
  $\xi$ & A bitstring, usually an address. \\
  $\text{dist}(x, y)$ & Hamming distance between $x$ and $y$. \\
  $\text{d}(x, y)$ & Same as $\text{dist}(x, y)$.
\end{tabular}\\

@@

\chapter{Results (iii): Loss of neurons}
In SDM, the data is written distributed among millions of hard-locations, which theoretically gives SDM robustness against loss of neurons. In other words, SDM should keep convering correctly even when some neurons are dead. The question is: how robust it really is? How many neurons may die before it starts to forget things. These questions have never been addresses before.

Looking for answers to these questions, we run simulations in which we kept killing some neurons and checking whether SDM remained converging to a given bitstring or not. In these simulations, 10,000 random bitstrings were written to a 1,000-bit SDM with 1,000,000 hard-locations, and we choose one of them as our target. As the bitstrings were all written exactly once, we may generalize the results. The code is available in the ``Reseting hard-locations'' notebook \citep{sdmframework}.

As neurons are hard-locations in SDM, when we say that a neuron has been killed, we mean that its counters have been zeroed and a new random bitstring address has been assigned. During our simulations, no other bitstring has been written after the 10,000. Consequently, as their counters willl remain zeroed, it is exactly like ignoring the dead hard-locations in the subsequent reading operations.

@@

In Figure \ref{fig:sdm-neuron-death-200k}, we can notice that SDM is robust up to 200,000 neuron deaths which is 20\% of all hard-locations. Its robustness is astonishing.  In fact, SDM begins to be significantly affected by loss of neurons after 600,000 neuron deaths (Figure \ref{fig:sdm-neuron-death-1m}), and obviously forgets everything when all neurons are dead.

@@

We run exactly the same simulation for a 256-bit SDM with 1,000,000 hard-locations. The results were even more surprising, as the 256-bit SDM seems to be more robust to loss of neurons than the 1,000-bit SDM (see Figure \ref{fig:sdm-neuron-death-256bits}). Notice that the loss of 50\% of neurons barely affected the 256-bit SDM which remained functional even facing an enormous loss of 90\% of neurons.

@@

Murilo observed that the models of Kanerva's read ($z=1$) and Chada's read ($z=0$) were simple variations of a generalized read with an exponent $z$, which suggests experimenting with different values. Mathematically, let $A$ be the set of the counters of the activated hardlocation, and $c_i$ be the counter of the $i$-th bit. Then,

@@

My advisor, Alexandre Linhares, has proposed another write operation: an information-theoretical weighted write. In it, the sum of the counter's value is weighted based on the distance between each hard-location's address and the reading address. The logic behind it is to vary the importance of each hard-location inside the circle.  It is only natural that one encodes an item in closer hard locations with a stronger signal, and a natural candidate for this signal function is the amount of information contained in the distance between the item and each hard location.  Closer hard locations have lower probabilies and therefore should encode more information.

@@

But we must consider that not all hardlocations are activated in each write operation, which changes our probability function. Thus, let $r$ be the access radius then:

@@

A hypercube with n dimensions can be divided by two hypercubes with $n-1$ dimensions. Is there an algorithm that separates the area of each hard-location in such a form that there exists a function mapping each bitstring in $\{0,1\}^n$ to the set of hard locations it `belongs to'?  Though this would break Kanerva's assumption of a randomly yet uniformly distributed set of hard locations --- for a perfectly symmetrical set of hard locations ---, there could be large performance gains if such a mapping function from a bitstring to its corresponding set of nearest hard locations exists.

>>>>

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. This model may be applied in all sort of applications because of its incredible ability to closely reflect the human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where one's fear comes from. One just feels it. We may interpret this situation as clues of the present --- a dark alley; a giant metropolitan area; people going on about their lives mostly indifferent from each other; constrained routes ahead and behind you; a general lack of activity; that very feeling that something isn't right... without knowing precisely what isn't right, or even what `right' means... etc. --- recalling past experiences from memory and thus generating the feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, one's brain involuntarily makes an analogy and recognizes the possibility of danger. This flexibility into mapping one situation in another is an important human feature which is hard to replicate in computers.

SDM has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. \citet{Linhares2011} has shown that SDM respects the limits of short-term memory discussed by \citet{Miller1995} and \citet{Cowan2011}. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though intriguing results have been achieved using SDM, it requires counter-productive, duplicate, effort from researchers to build on top of previous work.

@@

Our motivation was our own effort to run our models. As there is no reference implementation, we had to implement our own and run several simulations to ensure that our implementation was correct and bug free. Thus, we had to deviate from our main goal --- which was to test our hypothesis and explore the `idea space' --- and to focus on the implementation itself. Furthermore, new members of our research group had to go through different source codes developed by former members.

@@

\begin{tabular}{cp{\textwidth}}
  $n$ & Number of dimensions, i.e., $n=1,000$. \\
  $N$ & Size of the binary space, $|\{0, 1\}^n| = 2^n$. \\
  $N'$ & Number of hard locations samples from $\{0, 1\}^n$. Its typical value is 1,000,000, as suggested by \citet{Kanerva1988}. \\
  $H$ & Same as $N'$. \\
  $r$ & Access radius, i.e., when $n=1,000$ and $N'=1,000,000$, its typical value is $451$. This value is calculated to activate, on average, one thousandth of $N'$. \\
  $\eta$ & A bitstring, usually a datum. \\
  $\eta_x$ & A clue $x$ bits away from $\eta$, i.e., $\text{dist}(\eta, \eta_x) = x$. \\
  $\xi$ & A bitstring, usually an address. \\
  $\text{dist}(x, y)$ & Hamming distance between $x$ and $y$. \\
  $\text{d}(x, y)$ & Same as $\text{dist}(x, y)$.
\end{tabular}\\

@@

\chapter{Results (iii): Loss of neurons}
In SDM, the data is written distributed among millions of hard locations, which theoretically gives SDM robustness against loss of neurons. In other words, SDM should keep convering correctly even when some neurons are dead. The question is: how robust it really is? How many neurons may die before it starts to forget things. These questions have never been addresses before.

Looking for answers to these questions, we run simulations in which we kept killing some neurons and checking whether SDM remained converging to a given bitstring or not. In these simulations, 10,000 random bitstrings were written to a 1,000-bit SDM with 1,000,000 hard locations, and we choose one of them as our target. As the bitstrings were all written exactly once, we may generalize the results. The code is available in the ``Reseting hard locations'' notebook \citep{sdmframework}.

As neurons are hard locations in SDM, when we say that a neuron has been killed, we mean that its counters have been zeroed and a new random bitstring address has been assigned. During our simulations, no other bitstring has been written after the 10,000. Consequently, as their counters willl remain zeroed, it is exactly like ignoring the dead hard locations in the subsequent reading operations.

@@

In Figure \ref{fig:sdm-neuron-death-200k}, we can notice that SDM is robust up to 200,000 neuron deaths which is 20\% of all hard locations. Its robustness is astonishing.  In fact, SDM begins to be significantly affected by loss of neurons after 600,000 neuron deaths (Figure \ref{fig:sdm-neuron-death-1m}), and obviously forgets everything when all neurons are dead.

@@

We run exactly the same simulation for a 256-bit SDM with 1,000,000 hard locations. The results were even more surprising, as the 256-bit SDM seems to be more robust to loss of neurons than the 1,000-bit SDM (see Figure \ref{fig:sdm-neuron-death-256bits}). Notice that the loss of 50\% of neurons barely affected the 256-bit SDM which remained functional even facing an enormous loss of 90\% of neurons.

@@

Murilo observed that the models of Kanerva's read ($z=1$) and Chada's read ($z=0$) were simple variations of a generalized read with an exponent $z$, which suggests experimenting with different values. Mathematically, let $A$ be the set of the counters of the activated hard location, and $c_i$ be the counter of the $i$-th bit. Then,

@@

My advisor, Alexandre Linhares, has proposed another write operation: an information-theoretical weighted write. In it, the sum of the counter's value is weighted based on the distance between each hard location's address and the reading address. The logic behind it is to vary the importance of each hard location inside the circle.  It is only natural that one encodes an item in closer hard locations with a stronger signal, and a natural candidate for this signal function is the amount of information contained in the distance between the item and each hard location.  Closer hard locations have lower probabilies and therefore should encode more information.

@@

But we must consider that not all hard locations are activated in each write operation, which changes our probability function. Thus, let $r$ be the access radius then:

@@

A hypercube with n dimensions can be divided by two hypercubes with $n-1$ dimensions. Is there an algorithm that separates the area of each hard location in such a form that there exists a function mapping each bitstring in $\{0,1\}^n$ to the set of hard locations it `belongs to'?  Though this would break Kanerva's assumption of a randomly yet uniformly distributed set of hard locations --- for a perfectly symmetrical set of hard locations ---, there could be large performance gains if such a mapping function from a bitstring to its corresponding set of nearest hard locations exists.

### 2018-03-15 Chapters/sdm-applications.tex
Fix hard location (removing hard-location and hardlocation).

A simplified mathematical analysis would be: During the learning phase, $200$ images with 15\% of noise were written to SDM, so, the average distance between them and the clean image was 150 bits. Thus they shared, on average, 175 hard-locations with the clean image. In these 175 hard-locations, the counter's value for a black pixel mapped to bit 1 was $(1-0.15) \cdot 200 - 0.15 \cdot 200 = 140$. Finally, let's analyse the reading. When reading from a noisy image with 42\% of noise, the average distance between the noisy image and its clean image is 420, which means they share, on average, 6 hard-locations. As the average number of activated hard-locations is 1,072, the sum of their counters will be, on average, $Y = 6 \cdot 140 - \sum_{i=1}^{1072-6} X_i$, where $X_i$ is a Bernoulli trial with probability 0.5. Hence, $P(\text{black pixel}) = P(Y > 0) = P(6 \cdot 140 - \sum_{i=1}^{1072-6} X_i > 0) = P(\sum_{i=1}^{1066} X_i < 840) = 99.99\%$. But, when reading from a noisy image with 45\% of noise, the average number of hard-locations shared with the clean image is only 3. Thus the sum of the activated hardlocations' counters will be, on average, $3 \cdot 140 - \sum_{i=1}^{1072-3} X_i$, and $P(\text{black pixel}) = P(\sum_{i=1}^{1069} X_i < 420) = 1.28 \cdot 10^{-12}$. The probability abruptly drops from 100\% to 0\% when the noise goes from 42\% to 45\% (see Figure \ref{fig:filter-prob-right-pixel}). The analysis for while pixels is exactly the same, but with opposite signs. The code to calculate this probability is available in ``Noise filter - Math analysis'' notebook \citep{sdmframework}.

This is a simplified analysis because it does not take into consideration the hardlocations shared by the different letters. It works fine for our example because letter ``I'' and ``T'' are almost orthogonal and share, on average, only one hard location.

@@

If the intersection between images becomes too high, the noise filter stops working properly. We have confirmed it writing letters ``B'', ``C'', ``D'' and numeral ``8''. They share a great amount of hard-locations and our noise filter could not filter their noise correctly. The training sets can be seen in Figure \ref{fig:filter-training2} and the results in Figure \ref{fig:filter-results2}.

@@

The chunck through exclusive OR (XOR) works because of Theorem \ref{thm:xor-chuncks}, which says that chuncking the images with labels will generate, on average, orthogonal bitstrings. Thus, these orthogonal bitstrings will not interfere with each other because they share, on average, only one hardlocation.

>>>>

A simplified mathematical analysis would be: During the learning phase, $200$ images with 15\% of noise were written to SDM, so, the average distance between them and the clean image was 150 bits. Thus they shared, on average, 175 hard locations with the clean image. In these 175 hard locations, the counter's value for a black pixel mapped to bit 1 was $(1-0.15) \cdot 200 - 0.15 \cdot 200 = 140$. Finally, let's analyse the reading. When reading from a noisy image with 42\% of noise, the average distance between the noisy image and its clean image is 420, which means they share, on average, 6 hard locations. As the average number of activated hard locations is 1,072, the sum of their counters will be, on average, $Y = 6 \cdot 140 - \sum_{i=1}^{1072-6} X_i$, where $X_i$ is a Bernoulli trial with probability 0.5. Hence, $P(\text{black pixel}) = P(Y > 0) = P(6 \cdot 140 - \sum_{i=1}^{1072-6} X_i > 0) = P(\sum_{i=1}^{1066} X_i < 840) = 99.99\%$. But, when reading from a noisy image with 45\% of noise, the average number of hard locations shared with the clean image is only 3. Thus the sum of the activated hard locations' counters will be, on average, $3 \cdot 140 - \sum_{i=1}^{1072-3} X_i$, and $P(\text{black pixel}) = P(\sum_{i=1}^{1069} X_i < 420) = 1.28 \cdot 10^{-12}$. The probability abruptly drops from 100\% to 0\% when the noise goes from 42\% to 45\% (see Figure \ref{fig:filter-prob-right-pixel}). The analysis for while pixels is exactly the same, but with opposite signs. The code to calculate this probability is available in ``Noise filter - Math analysis'' notebook \citep{sdmframework}.

This is a simplified analysis because it does not take into consideration the hard locations shared by the different letters. It works fine for our example because letter ``I'' and ``T'' are almost orthogonal and share, on average, only one hard location.

@@

If the intersection between images becomes too high, the noise filter stops working properly. We have confirmed it writing letters ``B'', ``C'', ``D'' and numeral ``8''. They share a great amount of hard locations and our noise filter could not filter their noise correctly. The training sets can be seen in Figure \ref{fig:filter-training2} and the results in Figure \ref{fig:filter-results2}.

@@

The chunck through exclusive OR (XOR) works because of Theorem \ref{thm:xor-chuncks}, which says that chuncking the images with labels will generate, on average, orthogonal bitstrings. Thus, these orthogonal bitstrings will not interfere with each other because they share, on average, only one hard location.

### 2018-03-15 Chapters/sdm-architecture.tex
Fix hard location (removing hard-location and hardlocation).

We split the SDM memory in two parts: the hard-location addresses and the hard-location counters. Thus, the addresses (bitstrings) of the hard-locations are stored in one array, while their counters in another. This makes possible to create multiple SDMs using the same address space, which would save computational effort to scan a bitstring in all the SDMs --- since they share the same address space, the activated hard-locations will be the same in all of them. As the slowest part of reading and writing operations is scanning the address space, the performance benefits are significant.

@@

Let a SDM memory with $N$ dimensions and $H$ hard-locations. Then, in a 64-bit computer, the array of hard-location addresses will use $H \cdot 8 \cdot \lceil N/64 \rceil$ bytes of memory, and there will be $H \cdot N$ hard-location counters. For example, in a SDM memory with 1,000 dimensions and 1,000,000 hard-locations, using 32-bit integers for the counters, the array of addresses will use 122MB of memory and the counters will use 3.8 GB of memory.

@@

There is also the naive algorithm, which really counts the number of ones checking bit by bit. It is available only to testing purposes and should never be used.

@@

The scan for activated hard-locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle (and their distances). Then, each operation uses these pieces of information in a different way.

\subsection{Scanning for activated hard-locations}

Scanning for the activated hard-locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(H)$. The algorithm which seems to best fit in our problem consumes $O(H)$ space and runs in $O(\log^n(H))$ \citep{chazelle1988functional}, which is really slower than $O(H)$ when, for instance, $H=1,000,000$ and $n=1,000$. For a review of the range reporting algorithms, see \citet{chan2011orthogonal}.

@@

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computational geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(H))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hard-locations' addresses are randomly sampled from the $\{0, 1\}^n$ space. Although each addresses' bit itself splits the hardlocations in half, it does not split the search space in half since both halves still must be convered by the algorithm. For instance, let's say we have $n=1,000$ dimensions with $H=1,000,000$ hard-locations, and we are scanning within a circle with radius $r=451$, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. Hence, the search space has not been split in half because both halves have been covered (and one of them should have been skipped).

Finally, as our best approach is to scan through all hard-locations, we may distribute the scan into many tasks which will be executed independently. The tasks may be executed in different processes, threads, or even computers. They may also run in the CPU or in the GPU. In this case, we may take into account both the time required to distribute the tasks and the time to receive their results.

@@

A counter is an array of integers which stores the data of all hard-locations. So, the counter's array has $n \cdot H$ integers.

When two counters are added in a third counter, there may occur an overflow. It is not supposed to be a problem because, by default, each counter is a signed 32-bit integer that can store any number between -2,147,483,648 and 2,147,483,647, which means they will not overflow with less writes than $2^{31}-1$ divided by the average number of activated hard-locations. For instance, when $n=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes before being possible to a counter to overflow.  Note also that it would be more likely to saturate the memory before any overflow.

>>>>

We split the SDM memory in two parts: the hard-location addresses and the hard-location counters. Thus, the addresses (bitstrings) of the hard locations are stored in one array, while their counters in another. This makes possible to create multiple SDMs using the same address space, which would save computational effort to scan a bitstring in all the SDMs --- since they share the same address space, the activated hard locations will be the same in all of them. As the slowest part of reading and writing operations is scanning the address space, the performance benefits are significant.

@@

Let a SDM memory with $N$ dimensions and $H$ hard locations. Then, in a 64-bit computer, the array of hard-location addresses will use $H \cdot 8 \cdot \lceil N/64 \rceil$ bytes of memory, and there will be $H \cdot N$ hard-location counters. For example, in a SDM memory with 1,000 dimensions and 1,000,000 hard locations, using 32-bit integers for the counters, the array of addresses will use 122MB of memory and the counters will use 3.8 GB of memory.

@@

There is also the naive algorithm, which really counts the number of ones checking bit by bit. It is available only for testing purposes and should never be used.

@@

The scan for activated hard locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle (and their distances). Then, each operation uses these pieces of information in a different way.

\subsection{Scanning for activated hard locations}

Scanning for the activated hard locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(H)$. The algorithm which seems to best fit in our problem consumes $O(H)$ space and runs in $O(\log^n(H))$ \citep{chazelle1988functional}, which is really slower than $O(H)$ when, for instance, $H=1,000,000$ and $n=1,000$. For a review of the range reporting algorithms, see \citet{chan2011orthogonal}.

@@

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computational geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(H))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hard locations' addresses are randomly sampled from the $\{0, 1\}^n$ space. Although each addresses' bit itself splits the hard locations in half, it does not split the search space in half since both halves still must be covered by the algorithm. For instance, let's say we have $n=1,000$ dimensions with $H=1,000,000$ hard locations, and we are scanning within a circle with radius $r=451$, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. Hence, the search space has not been split in half because both halves have been covered (and one of them should have been skipped).

Finally, as our best approach is to scan through all hard locations, we may distribute the scan into many tasks which will be executed independently. The tasks may be executed in different processes, threads, or even computers. They may also run in the CPU or in the GPU. In this case, we may take into account both the time required to distribute the tasks and the time to receive their results.

@@

A counter is an array of integers which stores the data of all hard locations. So, the counter's array has $n \cdot H$ integers.

When two counters are added in a third counter, there may occur an overflow. It is not supposed to be a problem because, by default, each counter is a signed 32-bit integer that can store any number between -2,147,483,648 and 2,147,483,647, which means they will not overflow with less writes than $2^{31}-1$ divided by the average number of activated hard locations. For instance, when $n=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard locations is 1,000 and it would require at least one million writes before being possible to a counter to overflow.  Note also that it would be more likely to saturate the memory before any overflow.

### 2018-03-15 Chapters/sdm-critical-distance.tex
Fix hard location (removing hard-location and hardlocation).

\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard-location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{fig:sdm-10000w-table-7-2}}
\end{figure}

@@

\caption{Results generated by the framework similar from Kanerva's original Table 7.2. Here we have a 1,000 bit, 1,000,000 hard-location SDM with (a) exactly 100 random bitstrings written into it and (b) steps of 1,000 random bitstrings written into it.
\label{fig:sdm-100w-table-7-2}}
\end{figure}

@@

Going further in the analysis, we calculated the probability of missing a bit when reading from SDM. After all, that's how Kanerva has originally found the curve. To do this, we used the following equations from our previous work \citep{brogliato2014sparse}. Let $d$ be the distance to the target, $h$ be the number of hard-locations activated during read and write operations, $s$ be the number of total stored bitstrings, $H$ be the number of total hard-locations, $w$ be the number of times the target was written into SDM, $\theta$ be the total random bitstrings in all $h$ hard-locations activated by read operation, and $\phi(d)$ be the average number of shared hard-locations activated two bitstrings $d$ bits away.

@@

For details and the proof of this equation, see \citet{brogliato2014sparse}. Although Kanerva has found a formula for $\phi(d)$ through an unsolved integral, and \citet{de1995geometrical} have proposed another way to calculate $\phi(d)$, we have used our framework to estimate the values of $d$. In order to do that, we used a Monte Carlo approach, generating many pair of random bitstrings $d$ bits away from them and calculating the average number of shared hard-locations between them. The code is available in the ``Calculate critical distance'' notebook \citep{sdmframework}.

@@

\section{Critical distance 209}

>>>>

\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{fig:sdm-10000w-table-7-2}}
\end{figure}

@@

\caption{Results generated by the framework similar from Kanerva's original Table 7.2. Here we have a 1,000 bit, 1,000,000 hard location SDM with (a) exactly 100 random bitstrings written into it and (b) steps of 1,000 random bitstrings written into it.
\label{fig:sdm-100w-table-7-2}}
\end{figure}

@@

Going further in the analysis, we calculated the probability of missing a bit when reading from SDM. After all, that's how Kanerva has originally found the curve. To do this, we used the following equations from our previous work \citep{brogliato2014sparse}. Let $d$ be the distance to the target, $h$ be the number of hard locations activated during read and write operations, $s$ be the number of total stored bitstrings, $H$ be the number of total hard locations, $w$ be the number of times the target was written into SDM, $\theta$ be the total random bitstrings in all $h$ hard locations activated by read operation, and $\phi(d)$ be the average number of shared hard locations activated two bitstrings $d$ bits away.

@@

For details and the proof of this equation, see \citet{brogliato2014sparse}. Although Kanerva has found a formula for $\phi(d)$ through an unsolved integral, and \citet{de1995geometrical} have proposed another way to calculate $\phi(d)$, we have used our framework to estimate the values of $d$. In order to do that, we used a Monte Carlo approach, generating many pair of random bitstrings $d$ bits away from them and calculating the average number of shared hard locations between them. The code is available in the ``Calculate critical distance'' notebook \citep{sdmframework}.

@@

\section{Critical distance of 209}

### 2018-03-15 Chapters/sdm-performance.tex
Fix hard location (removing hard-location and hardlocation).

Each scan on a 1,000-bits SDM with 1,000,000 hard-locations executes $10^9$ bit compares through $10^9/64 = 15,625,000$ XORs and calls to the built-in popcount. So, 10,000 writes execute $10^{13}$ bit compares, while a 6-iterative reading executes $6 \cdot 10^{12}$ bit compares. The heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} executed $3.05 \cdot 10^{15}$ bit compares. For comparison, the number of seconds since Jesus's birth is 63,639,648,000 = $6.36 \cdot 10^{10}$. The number of people who have ever lived on earth is estimated to be $1,08 \cdot 10^{11}$. There are approximtely $1.8 \cdot 10^{9}$ websites in the internet. A modern laptop can increment a counter $5 \cdot 10^{8}$ times per second. Hence, a naive implementation of SDM may take several hours --- or days --- to simply write 10,000 random bitstrings.

@@

A total of 8 kernels have been developed for benchmarking in our framework: single\_scan0, single\_scan1, single\_scan2, single\_scan3, single\_scan4, single\_scan5, single\_scan5\_unroll, and single\_scan6. Each scan uses a different algorithm to do exactly the same thing: calculate which hard-locations are inside the circle of the given bitstring. They differ in how they split the work between work-groups and how they explore the GPU architecture to obtain performance gains.

>>>>

Each scan on a 1,000-bits SDM with 1,000,000 hard locations executes $10^9$ bit compares through $10^9/64 = 15,625,000$ XORs and calls to the built-in popcount. So, 10,000 writes execute $10^{13}$ bit compares, while a 6-iterative reading executes $6 \cdot 10^{12}$ bit compares. The heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} executed $3.05 \cdot 10^{15}$ bit compares. For comparison, the number of seconds since Jesus's birth is 63,639,648,000 = $6.36 \cdot 10^{10}$. The number of people who have ever lived on earth is estimated to be $1,08 \cdot 10^{11}$. There are approximtely $1.8 \cdot 10^{9}$ websites in the internet. A modern laptop can increment a counter $5 \cdot 10^{8}$ times per second. Hence, a naive implementation of SDM may take several hours --- or days --- to simply write 10,000 random bitstrings.

@@

A total of 8 kernels have been developed for benchmarking in our framework: single\_scan0, single\_scan1, single\_scan2, single\_scan3, single\_scan4, single\_scan5, single\_scan5\_unroll, and single\_scan6. Each scan uses a different algorithm to do exactly the same thing: calculate which hard locations are inside the circle of the given bitstring. They differ in how they split the work between work-groups and how they explore the GPU architecture to obtain performance gains.

### 2018-03-15 Chapters/sdm-validation.tex
Fix hard location (removing hard-location and hardlocation).

\section{Number of activated hard-locations}

In his seminal work, Kanerva proposed to use a sample of 1,000,000 hard-locations in a 1,000 bits SDM. He also proposed to activate only 1,000 of them, on average. He calculated that an access radius of $r=451$ would activate, on average, 0.00107185004892 of the whole space, or, in this case, 1,071.85 hard-locations.

We extended his results, calculating the distribution of the number of activated hard-locations. As each hard-location has probability $p=0.00107185004892$ of being activated, the probability of activating exactly $a$ out of $H$ hard-locations follows a binomial distribution with mean $\mu = pH$ and standard deviation $\sigma = \sqrt{Hp(p-1)}$. In this case, $\mu = 1071.85$ and $\sigma = 32.72$.

In order to validate our scan algorithm, we have run 10,000 scans from a random bitstring and counted the number of activated hard-locations. The code is available in the ``Number of activated hard-locations'' notebook \citep{sdmframework}.

@@

  \caption{Histogram of the number of activated hard-locations in 10,000 scans from a random bitstring. The curve in red is the theoretical normal distribution with $\mu = Hp$ and $\sigma = p(p-1)H$.}
  \label{fig:validation-activated-hls}
\end{figure}

Besides the number of activated hard-locations, we have also extended Kanerva's results to calculate the distribution of distances between the center of the circle and the activated hard-locations. Let $A$ be the set of activated hard-locations, $\xi$ be the center of the circle, and $r$ be the access radius, then:

@@

In order to check Equation \ref{eq:prob-d-inside-circle}, we have calculated the distances of the activated hard-locations to the center of 1,000 random circles. The code is available in the ``Distances of activated hard-locations'' notebook \citep{sdmframework}.

@@

\caption{Histogram of the distances of activated hard-locations to the center of the circles. The curve in red is the theoretical distribution of Equation \ref{eq:prob-d-inside-circle}
\label{fig:validation-distance-activated-hls}}
\end{figure}

@@

Kanerva has calculated the intersection of two circles according to the distance between their centers. The intersection is important to understand how SDM works, because it affects directly the critical distance. When $\eta_d$ is inside the critical distance, then it will converge to $\eta$. In fact, it converges because they share a sufficient amount of hard-locations, i.e., the intersection of the circle around $\eta_d$ and $\eta$ is enough to converge. For further information about the relation between the critical distance and the intersection, see \citet{brogliato2014sparse}.

@@

  \caption{Number of hard-locations in the intersection of circles around two bitstrings $x$ bits away.}
  \label{fig:validation-intersection}
\end{figure}

@@

Each reading operation was performed summing the counters of all activated hard-locations from all three memories. For instance, to read the sequence $<A, B, C>$, we have activated the hard-locations around $C$ in $\text{sdm}_1$, we have also activated the hard-locations around $D$ in $\text{sdm}_2$, and, finally, we have also activated the hard-locations around $A$ in $\text{sdm}_3$. After summing the counters of all those hard-locations, we evaluate the resulting bitstring just as in the original read operation.

>>>>

\section{Number of activated hard locations}

In his seminal work, Kanerva proposed to use a sample of 1,000,000 hard locations in a 1,000 bits SDM. He also proposed to activate only 1,000 of them, on average. He calculated that an access radius of $r=451$ would activate, on average, 0.00107185004892 of the whole space, or, in this case, 1,071.85 hard locations.

We extended his results, calculating the distribution of the number of activated hard locations. As each hard location has probability $p=0.00107185004892$ of being activated, the probability of activating exactly $a$ out of $H$ hard locations follows a binomial distribution with mean $\mu = pH$ and standard deviation $\sigma = \sqrt{Hp(p-1)}$. In this case, $\mu = 1071.85$ and $\sigma = 32.72$.

In order to validate our scan algorithm, we have run 10,000 scans from a random bitstring and counted the number of activated hard locations. The code is available in the ``Number of activated hard locations'' notebook \citep{sdmframework}.

@@

  \caption{Histogram of the number of activated hard locations in 10,000 scans from a random bitstring. The curve in red is the theoretical normal distribution with $\mu = Hp$ and $\sigma = p(p-1)H$.}
  \label{fig:validation-activated-hls}
\end{figure}

Besides the number of activated hard locations, we have also extended Kanerva's results to calculate the distribution of distances between the center of the circle and the activated hard locations. Let $A$ be the set of activated hard locations, $\xi$ be the center of the circle, and $r$ be the access radius, then:

@@

In order to check Equation \ref{eq:prob-d-inside-circle}, we have calculated the distances of the activated hard locations to the center of 1,000 random circles. The code is available in the ``Distances of activated hard locations'' notebook \citep{sdmframework}.

@@

\caption{Histogram of the distances of activated hard locations to the center of the circles. The curve in red is the theoretical distribution of Equation \ref{eq:prob-d-inside-circle}
\label{fig:validation-distance-activated-hls}}
\end{figure}

@@

Kanerva has calculated the intersection of two circles according to the distance between their centers. The intersection is important to understand how SDM works, because it affects directly the critical distance. When $\eta_d$ is inside the critical distance, then it will converge to $\eta$. In fact, it converges because they share a sufficient amount of hard locations, i.e., the intersection of the circle around $\eta_d$ and $\eta$ is enough to converge. For further information about the relation between the critical distance and the intersection, see \citet{brogliato2014sparse}.

@@

  \caption{Number of hard locations in the intersection of circles around two bitstrings $x$ bits away.}
  \label{fig:validation-intersection}
\end{figure}

@@

Each reading operation was performed summing the counters of all activated hard locations from all three memories. For instance, to read the sequence $<A, B, C>$, we have activated the hard locations around $C$ in $\text{sdm}_1$, we have also activated the hard locations around $D$ in $\text{sdm}_2$, and, finally, we have also activated the hard locations around $A$ in $\text{sdm}_3$. After summing the counters of all those hard locations, we evaluate the resulting bitstring just as in the original read operation.
